### 2025-07-25 paper/quarto_aaai/paper.tex
mnist figure untangled

\iffalse
%Example, Multiple Authors, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
% \title{My Publication Title --- Multiple Authors}
\author {
    % Authors
    First Author Name\textsuperscript{\rm 1},
    Second Author Name\textsuperscript{\rm 2},
    Third Author Name\textsuperscript{\rm 1}
}
\affiliations {
    % Affiliations
    \textsuperscript{\rm 1}Affiliation 1\\
    \textsuperscript{\rm 2}Affiliation 2\\
    firstAuthor@affiliation1.com, secondAuthor@affilation2.com, thirdAuthor@affiliation1.com
}
\fi
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\usepackage{placeins}
\usepackage{siunitx}
\sisetup{uncertainty-mode = separate}
\DeclareMathSizes{10}{9}{7}{6.5}
\usepackage{enumitem}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/mediabag/../../paper/figures/poc.pdf}}

@@

Average reductions in \(\text{IP}\) range from around 7\% for
\emph{MNIST} to almost 60\% for \emph{Circ}. For the real-world tabular
datasets they are around 12\% for \emph{CH} and \emph{Cred} and almost
25\% for \emph{GMSC}; for \emph{Adult} and \emph{OL} we find no
significant impact of CT on \(\text{IP}\). Reductions in \(\text{IP}^*\)
are even more substantial and generally statistically significant,
although the average degree of uncertainty is higher than for
\(\text{IP}\): reductions range from around 20\% (\emph{Moons}) to
almost 90\% (\emph{Circ}). The only negative findings are for OL and
MNIST, but they are not statistically significant. A qualitative
inspection of the counterfactuals in Figure~\ref{fig-mnist} (columns
2-5) suggests recognizable digits 1-4 for the model trained with CT
(bottom row), unlike the baseline (top row).

@@

@@

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_body.png}}

@@

\caption{\label{fig-mnist}Visual explanations for \emph{MNIST} for BL
(top) and CT (bottom). \textbf{Plausibility}: col.~1 is a random factual
0 (blue); cols. 2-5 are corresponding \emph{ECCCo} counterfactuals in
target classes 1 to 4. \textbf{Actionability}: cols. 6-10 show
integrated gradients averaged over test images in classes 5 to 9.}

@@

For synthetic datasets, we always protect the first feature; for all
real-world tabular datasets we could identify and protect an \emph{age}
variable; for \emph{MNIST}, we protect the five upper and lower pixel
rows of the full image. Statistically significant reductions in costs
overwhelmingly point in the expected positive direction reaching up to
around 60\% for \emph{GMSC}. Only in the case of \emph{Cred}, average
costs increase, likely because any potential benefits from protecting
the \emph{age} are outweighed by the increase in costs required for
greater plausibility. The findings for \emph{Adult} and \emph{MNIST} are
not significant. A qualitative inspection of the class-conditional
integrated gradients in Figure~\ref{fig-mnist} (columns 6-10) suggests
that CT still has the expected effect: the model (bottom) is insensitive
(blue) to the protected rows of pixels; details of this experiment are
reported in the supplementary appendix.

@@

As our results indicate, counterfactual training produces models that
are more explainable. Nonetheless, these advantages come at the cost of
two important limitations.

@@

State-of-the-art machine learning models are prone to learning complex
representations that cannot be interpreted by humans. Existing
explainability solutions cannot guarantee that explanations agree with
these learned representation. As a step towards addressing this
challenge, we introduce counterfactual training, a novel training regime
that integrates recent advances in contrastive learning, adversarial
robustness, and counterfactual explanations to incentivize
highly-explainable models. Through extensive experiments, we demonstrate
that CT satisfies this goal while preserving the predictive performance
and promoting robustness of models. Explanations generated from CT-based
models are both more plausible (compliant with the underlying
data-generating process) and more actionable (compliant with
user-specified mutability constraints), and thus meaningful to their
recipients. In turn, our work highlights the value of simultaneously
improving models and their explanations.

>>>>

\iffalse
%Example, Multiple Authors, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
% \title{My Publication Title --- Multiple Authors}
\author {
    % Authors
    First Author Name\textsuperscript{\rm 1},
    Second Author Name\textsuperscript{\rm 2},
    Third Author Name\textsuperscript{\rm 1}
}
\affiliations {
    % Affiliations
    \textsuperscript{\rm 1}Affiliation 1\\
    \textsuperscript{\rm 2}Affiliation 2\\
    firstAuthor@affiliation1.com, secondAuthor@affilation2.com, thirdAuthor@affiliation1.com
}
\fi
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\usepackage{placeins}
\usepackage{siunitx}
\sisetup{uncertainty-mode = separate}
\usepackage{enumitem}
\DeclareMathSizes{10}{9}{7}{6.5}
\usepackage{subcaption}

@@

\includegraphics[width=0.85\linewidth,height=\textheight,keepaspectratio]{paper_files/mediabag/../../paper/figures/poc.pdf}

@@

Average reductions in \(\text{IP}\) range from around 7\% for
\emph{MNIST} to almost 60\% for \emph{Circ}. For the real-world tabular
datasets they are around 12\% for \emph{CH} and \emph{Cred} and almost
25\% for \emph{GMSC}; for \emph{Adult} and \emph{OL} we find no
significant impact of CT on \(\text{IP}\). Reductions in \(\text{IP}^*\)
are even more substantial and generally statistically significant,
although the average degree of uncertainty is higher than for
\(\text{IP}\): reductions range from around 20\% (\emph{Moons}) to
almost 90\% (\emph{Circ}). The only negative findings are for \emph{OL}
and \emph{MNIST}, but they are insignificant. A qualitative inspection
of the counterfactuals in Figure~\ref{fig-mnist-ce} suggests
recognizable digits for the model trained with CT (bottom row), unlike
the baseline (top row).

@@

\begin{minipage}{\linewidth}

@@

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_ce.png}}

@@

\subcaption{\label{fig-mnist-ce}\emph{Plausibility}: faithful
counterfactuals (\emph{ECCCo}).}

\end{minipage}%
\newline
\begin{minipage}{\linewidth}

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_ig.png}}

\subcaption{\label{fig-mnist-ig}\emph{Actionability}: class-cond.
integrated gradients.}

\end{minipage}%

\caption{\label{fig-mnist}Visual explanations for \emph{MNIST} for BL
(top rows) and CT (bottom rows).}

@@

For synthetic datasets, we always protect the first feature; for all
real-world tabular datasets we could identify and protect an \emph{age}
variable; for \emph{MNIST}, we protect the five top and bottom pixel
rows of the full image. Statistically significant reductions in costs
overwhelmingly point in the positive direction reaching up to around
60\% for \emph{GMSC}. Only in the case of \emph{Cred}, average costs
increase, likely because any benefits from protecting \emph{age} are
outweighed by an increase in costs required for greater plausibility.
The findings for \emph{Adult} and \emph{MNIST} are insignificant. A
qualitative inspection of the class-conditional integrated gradients in
Figure~\ref{fig-mnist-ig} suggests that CT still has the expected
effect: the model (bottom) is insensitive (blue) to the protected rows
of pixels; details are reported in the appendix.

@@

As our results indicate, counterfactual training produces models that
are more explainable. Nonetheless, these advantages come with certain
limitations.

@@

State-of-the-art machine learning models are prone to learning complex
representations that cannot be interpreted by humans. Existing work on
counterfactual explanations has largely focused on designing tools to
generate plausible and actionable for any model. In this work, we
instead hold models accountable for delivering such explanations. We
introduce counterfactual training: a novel training regime that
integrates recent advances in contrastive learning, adversarial
robustness, and CE to incentivize highly-explainable models. Through
theoretical results and extensive experiments, we demonstrate that CT
satisfies this goal while promoting adversarial robustness of models.
Explanations generated from CT-based models are both more plausible
(compliant with the underlying data-generating process) and more
actionable (compliant with user-specified mutability constraints), and
thus meaningful to their recipients. In turn, our work highlights the
value of simultaneously improving models and their explanations.

### 2025-07-25 paper/sections/experiments.qmd
mnist figure untangled

Average reductions in $\text{IP}$ range from around 7% for *MNIST* to almost 60% for *Circ*. For the real-world tabular datasets they are around 12% for *CH* and *Cred* and almost 25% for *GMSC*; for *Adult* and *OL* we find no significant impact of CT on $\text{IP}$. Reductions in $\text{IP}^*$ are even more substantial and generally statistically significant, although the average degree of uncertainty is higher than for $\text{IP}$: reductions range from around 20% (*Moons*) to almost 90% (*Circ*). The only negative findings are for OL and MNIST, but they are not statistically significant. A qualitative inspection of the counterfactuals in @fig-mnist (columns 2-5) suggests recognizable digits 1-4 for the model trained with CT (bottom row), unlike the baseline (top row).

![Visual explanations for *MNIST* for BL (top) and CT (bottom). **Plausibility**: col. 1 is a random factual 0 (blue); cols. 2-5 are corresponding *ECCCo* counterfactuals in target classes 1 to 4. **Actionability**: cols. 6-10 show integrated gradients averaged over test images in classes 5 to 9.](/paper/figures/mnist_body.png){#fig-mnist}

@@

For synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five upper and lower pixel rows of the full image. Statistically significant reductions in costs overwhelmingly point in the expected positive direction reaching up to around 60% for *GMSC*. Only in the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility. The findings for *Adult* and *MNIST* are not significant. A qualitative inspection of the class-conditional integrated gradients in @fig-mnist (columns 6-10) suggests that CT still has the expected effect: the model (bottom) is insensitive (blue) to the protected rows of pixels; details of this experiment are reported in the supplementary appendix.

>>>>

Average reductions in $\text{IP}$ range from around 7% for *MNIST* to almost 60% for *Circ*. For the real-world tabular datasets they are around 12% for *CH* and *Cred* and almost 25% for *GMSC*; for *Adult* and *OL* we find no significant impact of CT on $\text{IP}$. Reductions in $\text{IP}^*$ are even more substantial and generally statistically significant, although the average degree of uncertainty is higher than for $\text{IP}$: reductions range from around 20% (*Moons*) to almost 90% (*Circ*). The only negative findings are for *OL* and *MNIST*, but they are insignificant. A qualitative inspection of the counterfactuals in @fig-mnist-ce suggests recognizable digits for the model trained with CT (bottom row), unlike the baseline (top row).

::: {#fig-mnist layout-nrow=2 layout-ncol=1}

![*Plausibility*: faithful counterfactuals (*ECCCo*).](/paper/figures/mnist_ce.png){#fig-mnist-ce}

![*Actionability*: class-cond. integrated gradients.](/paper/figures/mnist_ig.png){#fig-mnist-ig}

Visual explanations for *MNIST* for BL (top rows) and CT (bottom rows).

:::

@@

For synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five top and bottom pixel rows of the full image. Statistically significant reductions in costs overwhelmingly point in the positive direction reaching up to around 60% for *GMSC*. Only in the case of *Cred*, average costs increase, likely because any benefits from protecting *age* are outweighed by an increase in costs required for greater plausibility. The findings for *Adult* and *MNIST* are insignificant. A qualitative inspection of the class-conditional integrated gradients in @fig-mnist-ig suggests that CT still has the expected effect: the model (bottom) is insensitive (blue) to the protected rows of pixels; details are reported in the appendix.
